---
title: "MB5370-Techniques in Marine Science 1- Module .01 Workshop 4- Integrating AI into code"
Subject: The developers toolbox and AI
name: Karah James
output: html_document
date: "2026-05-27"
---
#Introduction: Using Ai to help data visualization. As well as ways to help filter data and challenging previous workshop knowledge by applying familiar data wragngling and organizing tactics to create plots. 

```{r warning=FALSE}
 library(tidyverse)
 library(palmerpenguins)
```

```{r message=FALSE, warning=FALSE}
# Create a ggplot scatter plot using the pengiuns dataset. 
# Plot bill length on the x-axis and bill depth on the y-axis.
# Color the points by species and use geom_point with size 3.
# Add a minimal theme and clean labels. 
ggplot(data = penguins) +
  geom_point(mapping = aes(x = bill_length_mm, y = bill_depth_mm, color = species), size = 3) +
  theme_minimal() +
  labs(title = "Bill Length vs Bill Depth of Penguins",
       x = "Bill Length (mm)",
       y = "Bill Depth (mm)",
       color = "Species")
```

```{r message=FALSE, warning=FALSE}
# Create a ggplot scatter plot using the pengiuns dataset. 
# Plot bill_length_mm on the y-axis and different species on the x-axis, faceted by island. 
# Use custom colors for each species and add a minimal theme.
ggplot(data = penguins) +
  geom_point(mapping = aes(x = species, y = bill_length_mm, color = species), size = 3) +
  facet_wrap(~ island) +
  scale_color_manual(values = c("Adelie" = "blue", "Chinstrap" = "green", "Gentoo" = "orange")) +
  theme_minimal() +
  labs(title = "Bill Length of Penguins by Species and Island",
       x = "Species",
       y = "Bill Length (mm)",
       color = "Species")
```
# 4.4 Installing Chattr(Conversational Workspace engine)
```{r message=FALSE, warning=FALSE}
library("ellmer")
library("chattr")
```
# 4.3 Conditinal logic- filtering data
```{r}
# Simple Example: Checking Sea Surface Temperature (SST)
sst <- 30.2

if (sst > 29.5) {
  print("Warning: Marine heatwave threshold exceeded!")
} else {
  print("SST remains within baseline parameters.")
}

```

```{r}
#the function "mutate" allows an argument or certain parameter to be executed when explained within the code.
# Sample coral data
coral_monitoring <- tibble(
  site = c("Site_A", "Site_B", "Site_C"),
  depth_m = c(12, 35, 8)
)
# As you can see below, we are telling the code to recognize "zone" and "depth," if a zone has a certain depth below 30m then its labeled as a shallow reef and if a zone is above 30m depth then its a deep reef.
# Classify depth using if_else
coral_monitoring <- coral_monitoring %>% 
  mutate(zone = if_else(depth_m > 30, "Deep Reef", "Shallow Reef"))
head(coral_monitoring)
```

```{r}
coral_monitoring <- coral_monitoring %>% 
  mutate(reef_category = case_when(
    depth_m < 10  ~ "Lagoon / Flats",
    depth_m <= 30 ~ "Crest / Slope",
    depth_m > 30  ~ "Mesophotic / Deep",
    TRUE          ~ "Unclassified" # Catch-all remainder
  ))
head(coral_monitoring)
```

```{r}
marine_stations<- tibble(
  salinity=c(5,28,32,12)
)
```

```{r}
marine_stations<-marine_stations %>%
  mutate(environmental_type = case_when(
    salinity < 15 ~ "Estuarine",
    salinity <= 30 ~ "Brackish",
    salinity > 30 ~ "Marine" ,
  ))
head(marine_stations)
```
# 4.4 Automation: Iterating efficiency
```{r}
# for-loop repeats code chunk for each element in a designated sequence
for (year in 2020:2024) {
  print(paste("Processing climate data for year:", year))
}
```

```{r}
#tracks real-world data- this case it reads 12 monthly turtle tracking devices
# Loop across index sequences
transect_lengths <- c(50, 100, 25, 75)

for (i in seq_along(transect_lengths)) {
  print(paste("Transect number", i, "measures", transect_lengths[i], "metres."))
}
```

```{r}
site_areas <- c(144, 400, 625)
results <- numeric(length(site_areas)) # Must build an empty container first

for (i in seq_along(site_areas)) {
  results[i] <- sqrt(site_areas[i])
}
head(site_areas)
```

```{r}
library(purrr)
# Map directly and define your expected output type explicitly
mapped_results <- map_dbl(site_areas, sqrt)

```

```{r}
# Iterate a summary function over split lists of data
iris %>% 
  split(.$Species) %>% 
  map(~summary(.x))

```
# 4.5 Writing custom functions
```{r}
# Function Definition
calculate_coral_mortality <- function(initial_count, surviving_count) {
  
  # Logic safety switch using our conditional tools!
  if (initial_count <= 0) {
    stop("Initial count must be greater than zero.")
  }
  
  mortality_rate <- (initial_count - surviving_count) / initial_count
  return(mortality_rate)
}

# Utilizing your custom function
calculate_coral_mortality(initial_count = 120, surviving_count = 84)
```

#Student Exercise #4: Custom Function Construction Task
```{r}
convert_temp_c_to_f<-function(temp_c) {
  if (temp_c < -273.15) {
    stop("Temperature cannot be below absolute zero.")
  }
  temp_f <- (temp_c * 9/5) + 32
  return(temp_f)
}
convert_temp_c_to_f(temp_c=-30)
```

```{r}
convert_temp_c_to_f<-function(temp_c) {
  if (temp_c < -273.15) {
    stop("Temperature cannot be below absolute zero.")
  }
  temp_f <- (temp_c * 9/5) + 32
  return(temp_f)
}
convert_temp_c_to_f(temp_c=30)
```








