---
title: "MB5370-Techniques in Marine Science 1- Module. 02-Exercise 2.12"
subject: Data Science in R
author: "Karah James"
output: html_document
date: "2026-06-04"
---

#Introduction- KeyStone species
Here we are wrangling raw data from the Ross River Estuary fish surveys and the goal is to produce clean data to general tables and figures to represent the relationship between water quality and predatory fish assemblages.
This data analyzes the relationship between water quality and predatory fish assemblages along Ross River Estuary gradient. 

```{r}
setwd("C:/Users/jd131798/Downloads/data_analysis/Data/")
```
# Housekeeping
```{r}
rm(list=ls())
```

# loading necessary packages
```{r warning=FALSE}
library(tidyverse) #ggplot, lubridate, 
library(readxl) # read excel files
```

# Load files for datasets
## Metadata
```{r}
metadata<-read.csv("C:/Users/jd131798/Downloads/data_analysis/Data/estuary_metadata (1).csv")|>
   mutate(
    site_name_clean=str_to_lower(site_name), 
    site_name_clean=str_replace_all(site_name_clean, pattern=" ", replacement="_"),
    site_name_clean=str_trim(site_name_clean)
  )
```

```{r}
#Phase 1: Ingestion and Decontamination
#main dataset-- bind data all four sites to one data frame(Excel Tab Iteration)
catch1<-read_excel("C:/Users/jd131798/Downloads/estuary_catch_log.xlsx", sheet="Lower_ross")
catch2<- read_excel("C:/Users/jd131798/Downloads/estuary_catch_log.xlsx", sheet="Mid_Ross ")
catch3<-read_excel("C:/Users/jd131798/Downloads/estuary_catch_log.xlsx", sheet="ROSS MOUTH")
catch4<-read_excel("C:/Users/jd131798/Downloads/estuary_catch_log.xlsx", sheet="Upper Ross")
catch_data_raw<-bind_rows(catch1,catch2,catch3,catch4)
```


```{r}
catch_data_raw
head(catch_data_raw)
glimpse(catch_data_raw)
str(catch_data_raw)

# need to fix the site name to all lowercase
# cleans species names to lowercase- String Standardization
catch_data<-catch_data_raw|>
 mutate(
    species_clean=str_to_lower(species),
    species_clean=str_replace_all(species_clean,pattern=" ", replacement="_"), 
    species_clean= str_trim(species_clean)
  )|>
  mutate(
    site_clean=str_to_lower(site), 
    site_clean=str_replace_all(site_clean, pattern=" ", replacement="_"),
    site_clean=str_trim(site_clean)
  )
```

```{r}
#Load Taxaonomic Dictionary-- to input into catch log for species scientific names
species_dict<-read.csv("C:/Users/jd131798/Downloads/species_dictionary.csv")
head(species_dict)

```

```{r}
# joining scientific name to species name in catch data final(Phase 2: Taxonomic translation)
catch_data_final<-catch_data|>
left_join(species_dict, by = join_by(species_clean==common_name))
```

#--------------# Sensor Data-Sonde- dataset

```{r}
#Load sensor Data
sensor_raw<- read.csv("C:/Users/jd131798/Downloads/estuary_sonde_data.csv")

head(sensor_raw)
glimpse(sensor_raw)
str(sensor_raw)
```

```{r}
# Hardware errors-Temporal Parsing-String Standardization
#Change the dates in the dataset to formal date-time objects, N/A turbidity #sensor errors
sensor_data<- sensor_raw|>
  mutate(
    datetime=dmy_hm(timestamp),
    turbidity= na_if(turbidity, -999),
    salinity=ifelse(salinity<0, NA, salinity)
  )
head(sensor_data)
```



##Phase 2: The Relational Architecture
####Notes:

```{r}
#sensor sonde dataset, summarize into daily mean calculations for each #site-String Standardization
sensor_daily<-
sensor_data|>
  mutate(
    date= floor_date(datetime, "day"),
    date= as_date(date)
  )|>
  group_by(site, date)|>
  summarise(
    mean_temp=mean(temperature, na.rm=TRUE), 
    mean_salinity=mean(salinity, na.rm=TRUE), 
    .groups="drop"
  )

```

```{r}
#Grand join-merging catch data, spatial metadata, water quality into one dataframe
##taxonomic translation- scientific names to the common fish names
##Zero-catch framework
catch_data_finalized<-catch_data_final|>
  left_join(metadata, by=join_by(site_clean==site_name_clean))|>
  left_join(sensor_daily, by= join_by(site_clean==site, date))|>
   complete(
    nesting(site_clean, date, zone, mean_temp, mean_salinity), 
    scientific_name
  )|>
  mutate(
    count=coalesce(count, 0))

head(catch_data_finalized)
```
#Phase 4: Statistical Exrtaction 
```{r}
#generating a summarize table
catch_data_finalized|>
  group_by(scientific_name, zone)|>
  summarise(n=n(),
    sem_salinity=sd(mean_salinity,na.rm=TRUE)/sqrt((n)),
    median_salinity=median(mean_salinity, na.rm=TRUE),
    .groups="drop"
  )
```
#Phase 5: Visual Communication 
```{r}
fish_stats<- catch_data_finalized|>
  group_by(scientific_name, zone)|>
  summarise(n=n(),
    mean_catch = (mean(count, na.rm=TRUE)),
    sem_catch = (sd(count, na.rm=TRUE)/sqrt((n))),
    .groups="drop"
  )
print(fish_stats)
```

```{r warning=FALSE}
#visualize fish abundance along different salinity gradients
fish_stats|>
  ggplot(aes(x=zone, y=mean_catch, fill=zone))+
  geom_point(data=catch_data_finalized, aes(x=zone, y=count),
             poition=position_jitter(width=0.2), alpha=0.8, size=1, inherit.aes=FALSE)+
  geom_point(position="dodge", size=2, pch=23)+
  
  geom_errorbar(aes(ymin=mean_catch-sem_catch,ymax=mean_catch + sem_catch),
                position=position_dodge(0.9), width=0.2)+
  facet_wrap(~scientific_name,scales="fixed")+
  theme_minimal()+
  theme(
    strip.text=element_text(face="italic", size=10, colour="grey20"),
    legend.position="none"
  )+
  labs(
    title= " Predatory Fish Assemblages across Estuarine Salinity Gradients",
    subtitle= " Ross River Estuary System, May 2026",
    x="Spatial Zone",
    y="Abundance(per day)", 
    fill="Spatial Zone"
  )
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```


