---
title: "MB5370-Techniques in MS- Exercise 2.12"
output: html_document
date: "2026-06-04"
---

#Introduction
Here we are wrangling raw data from the Ross River Estuary fish surveys and the goal is to produce clean data to general tables and figures to represent the relationship between water quality and predatory fish assemblages.

```{r}
getwd()
```
# Housekeeping
```{r}
rm(list=ls())
```

# Packages
```{r}
library(tidyverse) #ggplot, lubridate, 
library(readxl) # read excel files
```

# Load files
## Metadata
```{r}
metadata<-read.csv("C:/Users/jd131798/Downloads/estuary_metadata.csv")|>
   mutate(
    site_name_clean=str_to_lower(site_name), 
    site_name_clean=str_replace_all(site_name_clean, pattern=" ", replacement="_"),
    site_name_clean=str_trim(site_name_clean)
  )

```

```{r}
#Phase 1: Ingestion and Decontamination
#main dataset-- bind data all four sites to one data frame
catch1<-read_excel("C:/Users/jd131798/Downloads/estuary_catch_log.xlsx", sheet="Lower_ross")
catch2<- read_excel("C:/Users/jd131798/Downloads/estuary_catch_log.xlsx", sheet="Mid_Ross ")
catch3<-read_excel("C:/Users/jd131798/Downloads/estuary_catch_log.xlsx", sheet="ROSS MOUTH")
catch4<-read_excel("C:/Users/jd131798/Downloads/estuary_catch_log.xlsx", sheet="Upper Ross")
catch_data_raw<-bind_rows(catch1,catch2,catch3,catch4)
```


```{r}
catch_data_raw
head(catch_data_raw)
glimpse(catch_data_raw)
str(catch_data_raw)


# cleans species names to lowercase
catch_data<-catch_data_raw|>
 mutate(
    species_clean=str_to_lower(species),
    species_clean=str_replace_all(species_clean,pattern=" ", replacement="_"), 
    species_clean= str_trim(species_clean)
  )|>
  mutate(
    site_clean=str_to_lower(site), 
    site_clean=str_replace_all(site_clean, pattern=" ", replacement="_"),
    site_clean=str_trim(site_clean)
  )
# need to fix the site name to all lowercase
```

```{r}
#Load Taxaonomic Dictionary-- to input into catch log for species scientific names
species_dict<-read.csv("C:/Users/jd131798/Downloads/species_dictionary.csv")
head(species_dict)

```

```{r}
# joining scientific name to species name in catch data final(Phase 2: Taxonomic translation)
catch_data_final<-catch_data|>
left_join(species_dict, by = join_by(species_clean==common_name))
```

#--------------# Sensor Data-Sonde- dataset

```{r}
#Load sensor Data
sensor_raw<- read.csv("C:/Users/jd131798/Downloads/estuary_sonde_data.csv")
sensor_raw
head(sensor_raw)
glimpse(sensor_raw)
str(sensor_raw)
```

```{r}
# Hardware errors and Temporal Parsing
#Change the dates in the dataset to formal date-time objects, N/A turbidity sensor errors
sensor_data<- sensor_raw|>
  mutate(
    datetime=dmy_hm(timestamp),
    turbidity= na_if(turbidity, -999),
    salinity=ifelse(salinity<0, NA, salinity)
  )
head(sensor_data)
```



##Phase 2: The Relational Architecture
####Notes:

```{r}
#sensor sonde dataset, summarize into daily mean calculations for each site
sensor_daily<-
sensor_data|>
  mutate(
    date= floor_date(datetime, "day"),
    date= as_date(date)
  )|>
  group_by(site, date)|>
  summarise(
    mean_temp=mean(temperature, na.rm=TRUE), 
    mean_salinity=mean(salinity, na.rm=TRUE), 
    .groups="drop"
  )

```

```{r}
catch_data_finalized<-catch_data_final|>
  left_join(metadata, by=join_by(site_clean==site_name_clean))|>
  left_join(sensor_daily, by= join_by(site_clean==site, date))|>
 ##Phase 3: The Zero-Catch Framework
   complete(
    nesting(site_clean, date, zone, mean_temp, mean_salinity), 
    scientific_name
  )|>
  mutate(
    count=coalesce(count, 0))

head(catch_data_finalized)
```
#Phase 4: Statistical Exrtaction 
```{r}
catch_data_finalized|>
  group_by(scientific_name, zone)|>
  summarise(n=n(),
    sem_salinity=sd(mean_salinity,na.rm=TRUE)/sqrt((n)),
    median_salinity=median(mean_salinity, na.rm=TRUE),
    .groups="drop"
  )
```
#Phase 5: Visual Communication 
```{r}
fish_stats<- catch_data_finalized|>
  group_by(zone, scientific_name)|>
  summarize(
    mean_catch=mean(count),
    sem_catch=sd(count)/sqrt((n)),
    .groups="drop"
  )
```

```{r}
fish_stats|>
  ggplot(aes(x=zone, y=mean_catch, fill=zone))+
  geom_point(data=catch_data_finalized, aes(x=zone, y=count),
             poition=position_jitter(width=0.2), alpha=0.8, size=1, inherit.aes=FALSE)+
  geom_point(position="dodge", size=2, pch=23)+
  
  geom_errorbar(aes(ymin=mean_catch-sem_catch,ymax=mean_catch + sem_catch),
                position=position_dodge(0.9), width=0.2)+
  facet_wrap(~scientific_name,scales="fixed")+
  theme_minimal()+
  theme(
    strip.text=element_text(face="italic", size=10, colour="grey20"),
    legend.position="none"
  )+
  labs(
    x="Spatial Zone",
    y="Abundance(per day)", 
    fill="Spatial Zone"
  )
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```


